---
title       : Decoding Health News
subtitle    : Using the Leek score to help
author      : Patrick Barta, patrickbarta <at> patrickbarta.com
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [bootstrap, mathjax]            # {mathjax, quiz, bootstrap}
ext_widgets : {rCharts: libraries/nvd3}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## The problem

Every day, we are all inundated by various health claims in the media:

* Alzheimer's Patients Can Store Memories, Study Suggests
* 1 in 6 Seniors Takes Dangerous Combos of Meds, Supplements: Study
* Michelangelo Likely Had Arthritis, Medical Experts Say

We know that media sometimes sensationalizes things.

*How about a quick and dirty way to evaluate these health claims?*

--- .class #id 

## A Bayesian solution: The Leek score

* In 2014, Jeff Leek proposed a checklist of items to look for, which,
if positive, suggested that we should believe the study.
* However, just because a study has some positive features, that doesn't
mean that we should believe it uncritically.
* The approach is purely Bayesian: we start out with a prior probability
that we believe the study, then adjust that prior by using subjective
likelihood ratios which increase the prior odds when the checklist items are
true, and which decrease them when the items are false.
* Checklist items are things like whether the study was a randomized control trial.

--- .class #id

## Posterior vs. prior by number of positive criteria
```{r, echo=FALSE}
library(tidyr)
library(rCharts)
post <- function(prior,numPositive) {
  # handle special cases
  if ((prior == 0.) || (prior == 1.)) {
   return(prior) 
  }
  
  priorOdds <- prior/(1-prior) 
  lrt <- 2^(2*numPositive - 6)
  posteriorOdds <- priorOdds * lrt
  posterior <- posteriorOdds/(1+posteriorOdds)
  return(posterior)
}

# set of probabilities
p <- 0:100/100

df <- data.frame(prior=p, check.names=FALSE)
df$"0" <- sapply(df$prior,function(x) post(x,0))
df$"1" <- sapply(df$prior,function(x) post(x,1))
df$"2" <- sapply(df$prior,function(x) post(x,2))
df$"3" <- sapply(df$prior,function(x) post(x,3))
df$"4" <- sapply(df$prior,function(x) post(x,4))
df$"5" <- sapply(df$prior,function(x) post(x,5))
df$"6" <- sapply(df$prior,function(x) post(x,6))

# restack for lattice-style graphics
df <- df %>% gather(positive, posterior, 2:8)
```

In the chart below, the colored number labels refer to the number of checklist items which are true.

```{r nvd3plot3, results = 'asis', comment = NA, message = F, echo = F}
# chart
n1 <- nPlot(posterior ~ prior, data = df, type = 'lineChart', group="positive")
n1$chart(useInteractiveGuideline=TRUE)
n1$xAxis(axisLabel = "Prior probability")
n1$yAxis(axisLabel = "Posterior probability", width=40)
n1$show('inline')
```

Notice how really improbable things stay improbable, and really probable things stay probable. "Extraordinary claims require extraordinary proof."

---

## More information?

Leek's original article is at:

* http://fivethirtyeight.com/features/a-formula-for-decoding-health-news/

This article also contains more detail of Leek's Bayesian approach.

There is an interactive demo at:

* https://patrickbarta.shinyapps.io/LeekScore/

Code for app is on GitHub:

* https://github.com/patrickbarta/LeekScore

This presentation is at:

* http://patrickbarta.github.io/LeekScore

Source for presentation is on the gh-branch of the app source repo.
